POPLMark goes Concurrent!\\Challenge problems for process calculi and behavioural types

### 2023-10-16 challenges/linearity.tex
Backport my changes into challenges/.

\subsection{Challenge: Linearity and behavioral type systems}
\label{sec:challenge:linearity-beh-types}
This challenge formalises a proof that requires reasoning about linearity of channels.
Linearity is the notion that a channel must be used exactly once in a process.
This is necessary to prove properties about session type systems.
Linear reasoning is also necessary to formalise e.g.\ linear and affine types for the pi-calculus and cut elimination in linear logics.

@@

\begin{lemma}[Preservation for $\stackrel{a}{\equiv}$]
  \label{le:presequiva}
  If \( \sconga{P}{Q} \) then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By case analysis on the \TirName{Sc} rule applied:
  \begin{description}
  \item[\TirName {Par-Comm/Assoc}] By rearranging sub-derivations noting that
    order does not matter for linear contexts
  \item[\TirName{Par-Inact}]
   Right-to-left by strengthening~\ref{lemma:strenD}. Vice-versa,
    by picking $T$ to be $\Tend$ and weakening~\ref{lemma:weak} part 2.
  \item[\TirName{Res-Par}] By case analysis on $x : T$ being linear or $\tend$ and applying weakening and strengthening accordingly.
  \item[\TirName{Res-Inact:}] Use weakening~\ref{lemma:weak} part 2
  \item[\TirName{Res}] Noting that order does not matter
  \end{description}
\end{proof}

Now, following Sangiorgi \& Walker define as the compatible equivalence relation induced by $\sconga\cdot\cdot$, namely add to reflexivity, symmetry  and transitivity the following:
 \begin{mathpar}
   \inferrule[Cong]{\sconga P Q}{\scong {C[P]}{C[Q]}}
 \end{mathpar}

\begin{lemma}[Preservation for $\equiv$]
  \label{le:presequiv}
    If \( \scong{P}{Q} \) then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By induction on the structure of the derivation of \( \scong{P}{Q} \), with an inner induction of the structure of a process context.
  \begin{description}
  \item[\TirName {Refl}] Immediate
      \item[\TirName {Sym}] By IH.
      \item[\TirName {Trans}]  By two appeal to the IH
          \item[\TirName {Cong}] By induction on the structure of $C$. If the context is a hole, apply Lemma~\ref{le:presequiva}. In the step case, apply the IH: for example if $C$ has the form ${\Ppar C R}$, noting that ${\PBpar C R}[P]$ is equal to ${\Ppar {C[P]} R}$ we have \( \types{\Gamma;\Delta}{{\PBpar C R}[P]} \)  iff \( \types{\Gamma;\Delta}{{\PBpar C R}[Q]} \) by rule T-Par and the IH.
  \end{description}
\end{proof}

>>>>

\subsection{Challenge: Linearity and behavioral type systems}
\label{sec:challenge:linearity-beh-types}
This challenge formalises a proof that requires reasoning about linearity of channels.
Linearity is the notion that a channel must be used exactly once in a process
This is necessary to prove properties about session type systems, and the key
issue of this challenge is reasoning about the linearity of context splitting operations.
Linear reasoning is also necessary to formalise e.g.\ linear and affine types for the pi-calculus and cut elimination in linear logics.

@@

\begin{lemma}[Preservation for $\stackrel{a}{\equiv}$]
  \label{le:presequiva}
  If \( \sconga{P}{Q} \) then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By case analysis on the \TirName{Sc} rule applied:
  \begin{description}
  \item[\TirName {Par-Comm/Assoc}] By rearranging sub-derivations noting that
    order does not matter for linear contexts
  \item[\TirName{Par-Inact}]
   Right-to-left by strengthening~\ref{lemma:strenD}. Vice-versa,
    by picking $T$ to be $\Tend$ and weakening~\ref{lemma:weak} part 2.
  \item[\TirName{Res-Par}] By case analysis on $x : T$ being linear or $\Tend$ and applying weakening and strengthening accordingly.
  \item[\TirName{Res-Inact:}] Use weakening %~\ref{lemma:weak} part 2
  \item[\TirName{Res}] Noting that order does not matter
  \end{description}
\end{proof}

Now, following Sangiorgi \& Walker~\cite{picalcbook}, we give rules for the compatible
equivalence relation induced by $\sconga\cdot\cdot$, which we still
write as $\scong\cdot\cdot$: namely, add to reflexivity, symmetry and
transitivity the following condition:
 \begin{mathpar}
   \inferrule[Cong]{\sconga P Q}{\scong {C[P]}{C[Q]}}
 \end{mathpar}

\begin{lemma}[Preservation for $\equiv$]
  \label{le:presequiv}
    If \( \scong{P}{Q} \) then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By induction on the structure of the derivation of \( \scong{P}{Q} \), with an inner induction of the structure of a process context.
  \begin{description}
  \item[\TirName {Refl}] Immediate
      \item[\TirName {Sym}] By IH.
      \item[\TirName {Trans}]  By two appeal to the IH
      \item[\TirName {Cong}] By induction on the structure of $C$. If
        the context is a hole, apply Lemma~\ref{le:presequiva}. In the
        step case, apply the IH: for example if $C$ has the form
        ${\Ppar {C'} R}$, noting that ${\PBpar{ C'} R}[P]$ is equal to
        ${\Ppar {C'[P]} R}$ we have
        \( \types{\Gamma;\Delta}{{\PBpar {C'} R}[P]} \) iff
        \( \types{\Gamma;\Delta}{{\PBpar {C'} R}[Q]} \) by rule T-Par and
        the IH.
  \end{description}
\end{proof}

### 2023-10-16 challenges/scope-extrusion.tex
Remove mention of equality on names since it is obvious and confusing.
Backport my changes into challenges/.

This challenge formalises a proof that requires explicit scope extrusion.
Scope extrusion is the notion that a process can send restricted names to another process, as long as the restriction can safely be ``extruded'' (i.e.\ expanded) to include the receiving process.
This e.g.\ allows a process to set up a private connection by sending a restricted name to another process, then using this name for further communication.

@@

\subsubsection{Syntax}
In this section we assume that \emph{names}  can be compared for equality.
The syntax of processes is given by:
\[
\begin{array}{rcl}
  P,Q & :=& \Pend \quad\mid\quad \PBpar{P}{Q} \quad\mid\quad \Pout{x}{y}{P} \quad\mid\quad \Pin{x}{y}{P} \quad\mid\quad \Pres{x}{P}
\end{array}
\]

>>>>

This challenge formalises a proof that requires explicit scope extrusion.
Scope extrusion is the notion that a process can send restricted names to another process, as long as the restriction can safely be ``extruded'' (i.e.\ expanded) to include the receiving process.
This e.g.\ allows a process to set up a private connection by sending a restricted name to another process, then using this name for further communication.
The key issue of this challenge is reasoning about names that are ``in the process'' of being scope-extruded, which often presents difficulties for the mechanisation of binders.

@@

\subsubsection{Syntax}
The syntax of processes is given by:
\[
\begin{array}{rcl}
  P,Q & :=& \Pend \quad\mid\quad \PBpar{P}{Q} \quad\mid\quad \Pout{x}{y}{P} \quad\mid\quad \Pin{x}{y}{P} \quad\mid\quad \Pres{x}{P}
\end{array}
\]
